Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan

### 2025-03-07 analysis/paper/paper.qmd
work on discussion section and add references

---
title: "Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute: Academia Sinica
title-block-published: "Last updated"
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Colonial encounters between Europeans and Indigenous people often led to profound impacts on many Indigenous societies and cultures around the world. In northeastern Taiwan during the Late Iron Age, local Indigenous groups experienced intense foriegn contacts through due to regional trade networks in East Asia, which later expanded into a larger-scale trade after the arrival of Europeans in the 17th century. This research focuses on local subsistence through faunal remains to understand how Indigenous communities responded to the regional trade activities driven by the Europeans and the Han Chinese, particularly the high demand for natural resources, such as deer. By analyzing faunal remains from the upper cultural layer of Kiwulan (AD 1450-1950), a late Iron Age site in northeastern Taiwan, the local participation in deer hide trade and their adaptive strategies were discussed. The results suggest indigenous cultural persistence of deer consumption, along with increased reliance on suids during the European period as the demands for deer products intensified. This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European colonial territories, a perspective rarely discussed.

@@

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century was usually associated with long-distance trade, exploration, and colonization in many parts of the world, which often led to profound impacts on indigenous societies. Depending on particular geographic and historical circumstances, many regions experienced different forms of European colonialism and power relations [@Jordan2009]. Understanding the dynamics of these colonial processes has become a significant issue in archaeology because it provides critical insights into the complex outcomes of European colonization [@Gosden2004]. In East Asia, European colonization was mainly motivated by the acquisition of raw materials, natural resources, or luxury items for commercial purposes [@Prakash1997]. The new exchange systems and colonial institutions introduced by Europeans often affected local economic systems, social structures, and settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, many recent studies also suggest that indigenous groups did not just accept European colonization passively but actively responded to the arrival of Europeans and colonialism through resistance, negotiation, and cultural persistence and resilience, showing entangled colonial experiences [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998; @Noe2023].

@@

# European Encounters in the Northern Taiwan
<!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
In 1626, the Spanish East Indies Company (Indias Orientales Españolas) arrived in northern Taiwan and established the San Salvador colony at Heping Dao in Keelung, followed by Santo Domingo in Tamsui (@fig-KWL-map), to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp. 103-134; @Berrocal2018]. The neighboring area of the fort was inhavited by local indigenous groups, the Taparris and the Quimaurris, who had lived at the location of the fort before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. The relations between the Spanish and indigenous people might not involve extensive colonialism or intercultural domination, and the Spanish territorial control could be limited to the fort. Instead, the Spanish relied on the exchange of local goods and food surpluses with local indigenous people, according to written records [@Borao2001, SIT:170-171]. This corresponds with a higher proportion of indigenous and Chinese items with little European material found in San Salvador [@Berrocal2018], indicating cultural entanglement.

<!-- the Dutch domination, how and why colonial powers were able to exert control -->
In 1642, the Dutch defeated the Spanish and took over San Salvador and Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662. Compared to the Spanish, the Dutch were more strategic in consolidating their power by forming alliances with indigenous villages, implementing political and economic institutions, and conquering villages that were unwilling to comply [@Kang2016; @Andrade2007]. For example, the VOC governors used various colonial institutions to secure their authority, including the annual landdag gathering, taxation, tribute system, village lease system, census investigations, and the selection of indigenous representatives [@Kang2016; @Andrade1997; @Andrade2007]. Among the political institutions, the tribute system operated from 1642 to 1647 and the village lease system introduced in 1644 had great impacts on the local indigenous economy and environment. Based on the tribute system, each household in allied villages was required to pay rice or high-quality deerskin annually. The amount of deerskins collected from indigenous people in northern Taiwan once reached 7,000 pieces in 1646 [@Andrade2007]. Moreover, to increase deerskin exports to Japan, the Dutch implemented a hunting license system (1636-1645) that gave Han Chinese permission to hunt deer on indigenous lands during certain periods [@Heyns2001]. From 1633 to 1661, the average number of deerskins exported from Taiwan was about 67,000 per year [@Koo2011; @Cheng2017]. Those colonial activities accelerated the decline in the deer population and could lead to a harsh situation for the indigenous economy.

<!--variation of colonial powers across regions and indirect colonization at Kavalan-->
Although the Dutch power expanded across nearly all of Taiwan in the middle of the 17th century, the extent of their colonial dominance and control differed greatly among different regions. Colonial rules in regions far from the European forts were usually difficult and unsuccessful. Yilan in northeastern Taiwan is an example of minimal European presence and indirect colonization, with fewer imposed colonial rules for most of the time [@Kang2012]. Yilan is an alluvial plain, located about 70 kilometers from the European colony in Keelung, with mountains in between. The main inhabitants in Yilan were the local indigenous group, the Kavalan. Given its strategic location on the route to Torrobouan, where the Dutch believed gold mines existed, they were eager to bring this area under their control [@DZI2000; @Kang2016, p47]. In 1644, the Dutch invaded Yilan to incorporate the defeated Kavalan villages into their northern colonial territory and demanded tribute from the indigenous people [@Kang2016]. Instead of engaging directly with the Kavalan people, the Europeans relied heavily on the assistance of the Basay people active in northern Taiwan to coordinate and collect tributes in Yilan [@Kang2016]. This indirect approach allowed the maintenance of autonomy and flexibility for the Kavalan people, who were able to negotiate tribute requested by the VOC [@Kang2016].

<!-- the pre-existing trade network in norhtern Taiwan-->
Before the arrival of the Europeans in northern Taiwan, an inter-insular trade network had been established since the early 15th century, involving Chinese merchants and indigenous groups in Tamsui and Keelung [@Chen2005]. Chinese traders transported goods such as textiles, porcelains, beads, and metal items to exchange for raw materials from Taiwan, such as gold, sulfur, and animal hides and meats from the local Basay people [@Chen2005; @Andrade2007]. This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom during the 15th and 16th centuries and later to European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007].Yilan was included in this long-established trade network, where the Kavalan people exchanged products such as rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003]. This is evidenced by the large amount of imported foreign goods found across archaeological sites in Yilan, indicating frequent trade activities, especially around the 17th century [@TaiwanHistory2005]. Yilan experienced frequent trading activities driven by the pursuit of foreign goods even before the Europeans arrival, and these activities continued during the European presence in northern Taiwan, likely with some occasional pressure from European powers. This unique historical and spatial context makes Yilan an ideal location to study indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp. 31-78; @Trabert2017].

# Archaeological Context at Kiwulan
<!--introduction of Kiwulan and the population-->
Kiwulan, located in the northern part of Yilan (@fig-KWL-map), is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. As a salvage project, a large scale excavation was conducted at Kiwulan to recover information before it was destroyed. A total of 262 excavation units uncovered abundant artifacts and features, including middens, burials, and post-holes with wooden posts in situ. It suggests that Kiwulan is a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004]. The main inhabitants of Kiwulan were the Kavalan people, as indicated by local oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004]. According to a mid-17th-century Dutch census, there were about 40 indigenous villages with approximately 10,000 people, and Kiwulan (referred to as Kibannoran in Dutch records) was the largest village, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which led to the relocation of Kavalan people to eastern Taiwan [@Zhan1995].

<!--archaeological materials-->
The most common artifact was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan [@TaiwanHistory2005]. In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts, were largely found in both the residential area and burials. Due to no evidence of local production of beads or metallurgy, these items are believed to have been acquired through exchange, primarily with the Basay people active in northern Taiwan. This indicates that Kiwulan has continuously participated in regional trade networks closely associated with northern Taiwan since the 15th century and influenced by the long-distance trade during the 17th century [@Wang2007]. Moreover, the use of foreign goods appears to have cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021].

<!--archaeological remains-->
There were a wide range of animal bones and plant remains uncovered, reflecting the diversity of the local diet and the environment close to river and coastal area. The animals include deer, pigs, cows, birds, fish, shellfish, and plants mainly consisting of rice, peach, pandanus, and cucurbits [@Chen2004]. Rice farming seems to be common but limited to a household scale, according to historical records [@]. Shellfish consist of bivalves such as clams and oysters and univalves from rivers and brackish water environments, which indicates a reliance on aquatic and marine resources. Those archaeological remains correspond with tools related to subsistence found at Kiwulan, including iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities. According to those pieces of evidence, the subsistence practices at Kiwulan should be primarily reliant on rice farming, supplemented by hunting and fishing for wild resources. The resource use during the periods of external impacts still need further examination to better understands the changes over time.

@@

# Materials and Methods
To explore changes in dietary practices before and after the European arrival with associated long-distance trade, I examined faunal assemblages from 40 adjacent units at the residential area excavated during the first season (2001-2003) at Kiwulan [@Chen2007]. The remains considered in this study are from dwelling spaces and trash pits that could reflect daily dietary practices, whereas those from burials that might relate to ritual activities are excluded. These 40 units were selected because of their location that were believed to be the center of the settlement with thick archaeological deposits and relatively intact archaeological contexts. Also, the spatial continuity can provides a more consistent framework for understanding food practices over time on a settlement scale. The same excavation and recovery methods applied at the site allow comparison between assemblages from different temporal sequences.

To construct a reliable fine-grained chronological sequence, archaeological deposits were divided into six cultural layers (CL6-CL1) spanning from the 14th century to the 19th century based on excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators. These diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks [@Wang2011; @Hsieh2009]. Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could offer a relatively robust framework for comparing archaeological records before and after the European influence. Among these six temporal layers, the fourth layer can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second layer corresponds to the 19th century, associated with the large influx of Chinese migrants in Yilan.

Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences. Deer antlers without attaching to crania were excluded since they were less likely related to dietary practices. To obtain generalized pattern of resource use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of animal consumption [@Mckechnie2016; @Vanderwarker2010]. Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers. Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when necessary, such as for estimating the expected NISP at a site [@Grayson2004; @Lyman2012]. Different from the minimum number of individuals (MNI) that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of dietary practices by considering all fragmented bone assemblages. Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site with less impacts from sample size [@Grayson1984; @Lyman2008]. For comparison of identified taxa across classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004]. The normalized NISP is calculated by dividing NISP by the number of skeletal elements.

Given that deerskin was a primary commodity in local regional exchange since the 14th century and was in high demand during the Dutch colonization period [@Cheng2017], the frequencies of skeletal parts and surface modifications of deer bones were examined to assess butchery and transport strategies. This can shed light on how indigenous people utilize deer in response to high economic demand when involved in regional trade network. Skeletal parts were quantified using NISP and classified into groups based on meat yield and portions: meaty (e.g., upperlimb), mid-meaty (e.g., vertebrae), and less meaty parts (e.g., skull and lower limb) [@Madrigal2004; @Munro2004]. Teeth without associated cranial elements were excluded here to reduce overrepresentation due to their higher preservation and quantities compared to other parts. While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain [@Madrigal2002]. Thus, less meaty elements were considered as secondary source of nutrition in the interpretation. Similarly, the degree of fragmentation could also imply the extraction of marrow or grease, which is calculated by dividing the Minimal Number of Elements (MNE) by NISP. Bone surface modifications relative to human butchering and cooking activities, such as cut marks, fracturing, and burning damage, were recorded, as well as natural marks such as weathering and root etching for understanding taphonomic effects. Cut marks on major limb joints, such as shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were further quantified to explore any changes in butchering patterns between temporal sequences. Moreover, cut marks on lower limbs relative to skinning processes were examined and compared. Those marks are charactersitic of circular and straight incisions left by the series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].

@@

fauna_combined_context_all <-
  rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
  filter(!`部位/名稱` %in% c("角","犄角","角?")) %>%  #not associated with diet
  filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
  mutate(`重量(g)` = as.numeric(`重量(g)`)) %>%
  group_by(taxa) %>%
  mutate(`Weight (g)` = sum(`重量(g)`, na.rm = T)) %>%
  ungroup() %>%
  mutate(period = factor(period, levels = c("CL1","CL2","CL3","CL4","CL5","CL6"), order = T)) %>%
  mutate(category = case_when(str_detect(animal, "sika")|str_detect(animal, "sambar")|
                              str_detect(animal, "e deer")~ "deer", str_detect(animal, "pig")|
                                str_detect(animal, "boar") ~ "suid", str_detect(animal, "fish") ~ "fish",
                              str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "cattle",
                              animal == "aves" ~ "bird", TRUE ~ animal)) %>%
  mutate(class = case_when(str_detect(taxa, "muntjac")|str_detect(taxa, "ammal")|str_detect(taxa, "erv")|
                           str_detect(taxa, "Sus")|str_detect(taxa, "Rattus")|str_detect(taxa, "Rusa")|
                           str_detect(taxa, "Bos")|str_detect(taxa, "Bubalus")~ "mammal",
                           str_detect(category, "fish") ~ "fish", str_detect(category, "bird") ~ "bird",
                           TRUE ~ "reptile")) %>%
  mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
  mutate(category = fct_relevel(category, "deer","suid","muntjac","cattle","rat","bird","fish","turtle")) %>%
  mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"), "yes", "no"),
         fractures = ifelse(str_detect(`人為痕跡`, "削")|str_detect(`人為痕跡`, "砍")|
                            str_detect(`人為痕跡`, "折")|str_detect(`人為痕跡`, "敲"), "yes", "no"),
         burning =  ifelse(str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒"), "yes", "no")) %>%
  filter(is.na(refitted)) # for NISP

>>>>

---
title: "Indigenous Subsistence and Colonial Encounters: A Zooarchaeological Perspective from Kiwulan in Northeastern Taiwan"
author:
  - Li-Ying Wang:
      correspondence: "yes"
      email: liyingw15@gmail.com
      orcid: 0000-0001-9060-2125
      institute: Academia Sinica
title-block-published: "Last updated"
date: now
date-format: long
format: 
  docx:
    reference-doc: "../templates/template.docx" # Insert path for the DOCX file
execute:
  echo: true
  warning: false
  message: false
  comment: "#>"
  fig-path: "../figures/"
  fig-dpi: 600
filters:
  - ../templates/scholarly-metadata.lua
  - ../templates/author-info-blocks.lua
  - ../templates/pagebreak.lua
bibliography: references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  Colonial encounters between Europeans and Indigenous peoples often had profound impacts on Indigenous societies and cultures in many parts of the world. In northeastern Taiwan during the Late Iron Age, local Indigenous groups experienced intense foriegn contact through regional trade networks in East Asia, which expanded further with the arrival of Europeans in the 17th century. This study focuses on local subsistence through faunal remains to explore how Indigenous communities responded to trade activities driven by Europeans and Han Chinese, particularly the high demand for natural resources such as deer. By analyzing faunal remains from the upper component of Kiwulan (AD 1350-1850), a late Iron Age site in northeastern Taiwan, this research dicusses the Indigenous participation in the deerskin trade and their adaptive strategies. The results suggest cultural persistence in deer consumption, along with increased reliance on suids during the European period as demand for deer products intensified. This paper also provides insights into colonial experiences from Indigenous settlements at the periphery of European colonial territories, a perspective rarely explored.

@@

# Introduction
<!-- introduction about European expansion -->
European expansion since the 15th century was closely tied to long-distance trade, exploration, and colonization, which often had profound impacts on Indigenous societies. The impacts also varied across regions depending on specific geographic and historical circumstances, resulting in different forms of European colonization and power relations [@Jordan2009]. Understanding the diverse and dynamic colonial processes has become a key focus in archaeology, because it provides critical insights into the complex consequences of European expansion [@Gosden2004]. In East Asia, European colonization was mostly driven by the pursuit of raw materials, natural resources, or luxury goods for commercial purposes [@Prakash1997]. The new exchange systems and colonial institutions introduced by Europeans profoundly influenced local economies, social structures, or settlement patterns [@Christie1996; @Junker2004; @Lape2002]. However, recent studies emphasize that Indigenous groups were not just passively accept European colonization; instead, they actively responded through resistance, negotiation, and cultural persistence or resilience, leading to complex and entangled colonial experiences [@Torrence2003; @Silliman2005; @Mitchell2010; @Jordan2009; @Alexander1998; @Noe2023].

@@

# The European colonization period and regional trade network in Northern Taiwan
<!-- Spanish colonization and the power relations in particular spatial and temporal contexts-->
In 1626, the Spanish (Spanish East Indies) established the first colony of San Salvador at Heping Dao, and later Santo Domingo in Tamsui in northern Taiwan (@fig-KWL-map), to counter the threat of the VOC (Vereenigde Oost-Indische Compagnie; Dutch East India Company) in the Asian market, which had established a colony in southern Taiwan two years earlier [@Borao2009, p.41; @Andrade2007]. Although the Spanish intended to build a colonial city, San Salvador remains its early development during their colonization, primarily consisting of a fortress and a church [@Borao2009, pp. 103-134; @Berrocal2018]. The neighboring area of the fortress was inhabited by two Indigenous groups, the Taparris and the Quimaurris, who had lived at the fort location before being expelled by the Spanish [@Borao2001, SIT:165, p.81]. The relations between the Spanish and Indigenous people might not involve extensive colonialism or intercultural domination, with limited territorial control only to the fort. Also, the Spanish relied on the exchange of local goods and food surpluses with Indigenous people, according to written records [@Borao2001, SIT:170-171]. This corresponds with a higher proportion of Indigenous and Chinese items with little European materials found in San Salvador [@Berrocal2018], indicating cultural entanglement.

<!-- the Dutch domination, how and why colonial powers were able to exert control -->
In 1642, the Dutch defeated the Spanish and took over San Salvador and Santo Domingo, expanding their colonial power to northern and eastern Taiwan until 1662. Compared to the Spanish, the Dutch was more strategic in consolidating their power by establishing political and economic institutions, forming alliances with Indigenous communities, and conquering resistant communities [@Kang2016; @Andrade2007]. For examples, the colonial institutions to Indigenous people include taxation, tribute system, village-leasehold system, census investigations, annual gatherings (Dutch: Landdag), and Indigenous representative from each aliened village [@Kang2016; @Andrade1997; @Andrade2007]. Among those institutions, the tribute system operated from 1642 to 1647 and the village-leasehold system introduced in 1644 had great impacts on Indigenous economy and the environment they lived in. Based on the tribute system, each household in allied communities was required to pay rice or deer products annually, especially deerskin. The amount of deerskin collected from Indigenous communities in northern Taiwan once reached 7,000 pieces in 1646 [@Andrade2007]. Moreover, to increase deerskin exports to Japan, the Dutch implemented a hunting license system (1636-1645) that gave Han Chinese permission to hunt deer on Indigenous lands during certain periods [@Heyns2001]. From 1633 to 1661, the average number of deerskins exported from Taiwan was about 67,000 per year [@Koo2011; @Cheng2017]. Those colonial activities might affect the economy of indigenous people since deer was an important resource in prehistoric Taiwan [@Chen2000animal].

<!--variation of colonial powers across regions and indirect colonization at Kavalan-->
Although the Dutch power expanded across nearly whole Taiwan island in the mid-17th century, the extent of their colonial dominance differed greatly from region to region. Colonial power over regions far from the European fortress were usually difficult and unsuccessful. Yilan in northeastern Taiwan is an example of minimal European dominance or indirect colonization for most of the time [@Kang2012]. As an plain located around 70 kilometers away from the European colonies with mountains in between, Yilan was hardly accessed by land but mostly from sea. The main Indigenous group in Yilan was the Kavalan. Given its strategic location on the route to Torrobouan, a place believed having gold mines, the Dutch was eager to integrate Yilan under their control [@DZI2000; @Kang2016, p47]. In 1644, the Dutch defeated several Kavalan communities and incorporated them into their colonial territory [@Kang2016]. Instead of engaging directly with the Kavalan people, the Dutch relied mostly on the assistance of the mediate, the Basay people active in northern Taiwan [@Kang2003], to coordinate and collect tributes from Yilan. This indirect approach allowed the maintenance of autonomy of the Kavalan people, and was even able to negotiate tributes requested by the VOC [@Kang2016].

<!-- the pre-existing trade network in norththern Taiwan-->
Before the arrival of the Europeans in northern Taiwan, an inter-insular trade network had been established since the 14th century, involving Chinese merchants and indigenous groups in Tamsui and Keelung [@Chen2005]. Chinese traders transported goods such as textiles, porcelains, beads, and metal items to exchange for raw materials from Taiwan, such as gold, sulfur, and animal hides and meats from the Basay people [@Chen2005; @Andrade2007]. This regional trade network was further connected to maritime trade between Ming China and the Ryukyu Kingdom during the 15th and 16th centuries and later to European long-distance trade in the 17th century [@Chen2005; @Von2019; @Wang2007]. Yilan was included in this long-established trade network, where the Kavalan people exchanged products such as rice and deerskins for foreign goods and handicrafts from the Basay people [@Kang2003]. This is evidenced by the large amount of imported foreign goods found across archaeological sites in Yilan, indicating frequent trade activities, especially around the 17th century [@TaiwanHistory2005]. This trade network was then intensified during the European presence, and the participation of Kavalan in this network was driven by the pursuit of foreign goods, and also likely colonial pressure from European powers. This unique historical and spatial context makes Yilan an ideal location to study Indigenous reactions to foreign influences and the associated trade networks from a long-term perspective in a pericolonial context [@Acabado2017; @Kang2016, pp. 31-78; @Trabert2017].

# Archaeological Context at Kiwulan
<!--introduction of Kiwulan and the population-->
Kiwulan, located in the northern part of Yilan (@fig-KWL-map), is a multi-component site, where the upper component spans from the late Iron Age to the historical periods (AD 1350-1850) [@Chen2007]. A large scale salvage excavation was conducted at Kiwulan that exposed a total of 262 excavation units (4 * 4 m), revealing abundant artifacts and numerous features, including middens, burials, ditches, and post-holes with wooden posts in situ. The excavtion suggests that Kiwulan was a well-developed settlement with longhouses (approximately 40 x 80 m) situated side by side and burials located to the north of the residential area [@Chen2004; @Chiu2004]. The main inhabitants of Kiwulan were the Kavalan people based on oral history and the similarities between Kavalan material culture recorded in historical records and archaeological artifacts [@Chen2004]. Kavalan people is the main Indigenous group in Yilan during the late Iron Age and the early modern period. According to a Dutch census conduced in the mid-17th century, there were about 40 Kavalan settlements with approximately 10,000 people in Yilan. Kiwulan, written as "Kibannoran" in Dutch records, was the largest settlement, with 160 households and 711 people [@Nakamura1938]. In the early 19th century, the population of Kiwulan dramatically decreased to fewer than 100 people due to the population pressure of Han Chinese migrants, which eventrually led to the relocation of Kavalan people to Hualien in eastern Taiwan [@Zhan1995].

<!--archaeological materials-->
The most common artifact found at Kiwulan was locally made pottery with geometric patterns as surface treatments, which was also found at many simultaneous sites in Yilan [@TaiwanHistory2005]. In addition to pottery, a variety of foreign goods, such as porcelains, stonewares, glass beads, carnelian beads, and metal artifacts and ornaments, were largely found in both the residential area and burials. Due to no evidence of metallurgy and local production of beads, these items are believed to have been acquired through exchange, primarily with the Basay people active in northern Taiwan. This indicates that Kiwulan has continuously participated in regional trade networks closely associated with northern Taiwan since the 14th century and influenced by the long-distance trade during the 17th century [@Wang2007]. Moreover, the use of foreign goods appears to have cultural significance, and previous studies suggested a possible increase in social differentiation due to the introduction of foreign goods [@Wang2020ornament; @Wang2021; @Cheng2008].

<!--archaeological remains-->
Abundant animal bones, shellfish, and plant remains were uncovered at Kiwulan, reflecting the diversity of natural resources in the plain and an environment close to estuary. The animal resource includes deer, suids, cattle, birds, fish, shellfish, and plants mainly consisting of rice, peach, pandanus, and cucurbits [@Chen2004]. Rice farming seems to be common but likely limited to a household scale rather than intensive agriculture according to historical records [@Chen1963]. Shellfish consist of bivalves such as clams and oysters and univalves from rivers and brackish water environments. Based on those archaeological remains and environments, Kiwulan residents should have practiced a mixed subsistence that includes rice farming, hunting, and fishing for wild resources. Artifacts associated with those subsistence practices were also found at Kiwulan, such as iron arrowheads for hunting, iron sickles for crop harvesting, and wooden paddles for water-related activities. However, the resource use across different time periods still need further investigations to better understands the subsistence adaptation relative to foreign presence and environmental changes.

@@

# Materials and Methods
The faunal assemblages from Kiwulan analyzed in this study were sampled from 40 adjacent units at the residential area excavated during the first season (2001-2003) [@Chen2007]. These 40 units were selected because of their location that were believed to be the major residential area with thick archaeological deposits and relatively intact archaeological contexts without disruption from modern construction. The spatial continuity can provides a consistent framework for understanding food practices over time on a settlement scale. Also, the same excavation and recovery methods applied at the site allow comparison between assemblages from different temporal sequences. I mainly focus on vertebrate species due to their economic values in the trade market. The contexts of faunal remains include dwelling spaces and middens that were associated with daily dietary practices, whereas those from burials that might relate to ritual activities are excluded.

To construct a reliable fine-grained chronological sequence, archaeological deposits were divided into six cultural layers (CL6-CL1) spanning from the 14th century to the 19th century based on excavation depth, stratigraphic sections, sediment color and texture, radiocarbon dates, and archaeological diagnostic materials that served as time indicators. The diagnostic materials include stoneware commonly found at European associated sites in early 17th-century Southeast Asia, such as Anping jars and martavan [@Sakai2012; @Ketel2011; @Dizon2016; @Hsieh1995; @Lee2021], pipes, and Chinese bricks relating to the 19th century [@Wang2011; @Hsieh2009]. Although precisely identifying the stratum representing the European arrival is challenging due to its short duration, the separation of strata into six layers based on the criteria mentioned above could still offer a relatively robust framework for comparing archaeological records before and after the foreign influence. Among these six temporal layers, the fourth cultural layer (CL4) can be associated with the 17th century, when the Europeans occupied northern Taiwan, while the second cultural layer (CL2) corresponds to the 19th century, associated with the large influx of Chinese migrants in Yilan.

Only the faunal assemblages identified to at least the family level were considered in this study for consistent comparison across time sequences. Deer antlers without attaching to cranium were excluded since they were less likely related to dietary practices. To obtain generalized pattern of animal use, measures of ubiquity and taxonomic abundance were applied. Ubiquity measures the frequency of specific taxa through their presence or absence in particular depositional contexts, suggesting the regularity of consumtion of that taxa [@Mckechnie2016; @Vanderwarker2010]. Here ubiquity values are expressed as the percentage of units with a given taxon to the total number of units per cultural layers. Faunal assemblages were primarily quantified using the Number of Identified Specimens (NISP) method, with the Minimum Number of Individuals (MNI) applied when necessary, such as for estimating the expected NISP at a site [@Grayson2004; @Lyman2012]. Different from MNI that determines the actual numbers of animals recovered from remains, NISP offers a broad overview of remains by considering all fragmented bone assemblages. Despite the criticism of overestimation the number of animals, NISP captures better variation in species abundance on a site scale with less impacts from sample size [@Grayson1984; @Lyman2008]. For comparison across animal classes, the normalized NISP were used to avoid potential biases in identification due to differences in the preservation or recovery of skeletal elements [@Grayson2004]. The normalized NISP is calculated by dividing NISP by the number of skeletal elements.

Given that deerskin was a primary commodity in local regional exchange and in high demand during the Dutch colonization period [@Cheng2017], the frequencies of skeletal parts and surface modifications of deer bones were examined to assess butchery and transport strategies. This can shed light on how indigenous people utilize deer in response to high economic demand when involved in regional trade network. Skeletal parts were quantified using NISP and classified into groups based on meat yields and portions: meaty (e.g., upperlimb), mid-meaty (e.g., vertebrae), and less meaty parts (e.g., skull and lower limb) [@Madrigal2004; @Munro2004]. Teeth without associated cranial elements were excluded to reduce overrepresentation due to their much higher preservation rate and quantities compared to other parts. While lower limbs and skull bones have minimal meat, they are often valued in many cultures for marrow and soft tissues, such as tongue and brain [@Madrigal2002]. Thus, less meaty elements were considered as additional source of nutrition in the interpretation. Similarly, the degree of fragmentation could also imply the extraction of marrow or grease [@Lupo1997], which is calculated by dividing the Minimal Number of Elements (MNE) by NISP.

Deer bone surface modifications relative to human butchering and cooking practices, such as cut marks, fracturing, and burning damage, were recorded, as well as natural marks such as weathering and root etching for understanding taphonomic effects. Cut marks on major limb joints, such as shoulder, elbow, wrist, hip, knee, and ankle [@Lyman2005], were quantified to explore butchering patterns over time. Moreover, cut marks on lower limbs relative to skinning processes were examined and compared. Those marks are characteristic of circular and straight incisions left by the series of cuts for acquiring intact animal hides [@Costamagno2019; @Isaakidou2007].

The data and R code used for these analyses and visualizations are openly available at xxx for research reproducibility and transparency (@Marwick2017; @MarwickandWang2022).

@@

fauna_combined_context_all <-
  rbind(fauna_sam, fauna_H_sam) %>% # combine the two datasets
  filter(!`部位/名稱` %in% c("角","犄角","角?")) %>%  #not associated with diet
  filter(!`部位/名稱`== "角基部"|!is.na(`部位/左右`)) %>% # associated with cranial parts
  mutate(`重量(g)` = as.numeric(`重量(g)`)) %>%
  group_by(taxa) %>%
  mutate(`Weight (g)` = sum(`重量(g)`, na.rm = T)) %>%
  ungroup() %>%
  mutate(period = factor(period, levels = c("CL1","CL2","CL3","CL4","CL5","CL6"), order = T)) %>%
  mutate(category = case_when(str_detect(animal, "sika")|str_detect(animal, "sambar")|
                              str_detect(animal, "e deer")~ "deer", str_detect(animal, "pig")|
                                str_detect(animal, "boar") ~ "suid", str_detect(animal, "fish") ~ "fish",
                              str_detect(animal, "cattle")|str_detect(animal, "buffalo")~ "bovine",
                              animal == "aves" ~ "bird", TRUE ~ animal)) %>%
  mutate(class = case_when(str_detect(taxa, "muntjac")|str_detect(taxa, "ammal")|str_detect(taxa, "erv")|
                           str_detect(taxa, "Sus")|str_detect(taxa, "Rattus")|str_detect(taxa, "Rusa")|
                           str_detect(taxa, "Bos")|str_detect(taxa, "Bubalus")~ "mammal",
                           str_detect(category, "fish") ~ "fish", str_detect(category, "bird") ~ "bird",
                           TRUE ~ "reptile")) %>%
  mutate(`部位/左右` = case_when(`部位/左右` == "Ｌ" ~ "L", `部位/左右` == "Ｒ" ~ "R", TRUE ~ `部位/左右`)) %>%
  mutate(category = fct_relevel(category, "deer","suid","muntjac","bovine","rat","bird","fish","turtle")) %>%
  mutate(cutmarks = ifelse(str_detect(`人為痕跡`, "切"), "yes", "no"),
         fractures = ifelse(str_detect(`人為痕跡`, "削")|str_detect(`人為痕跡`, "砍")|
                            str_detect(`人為痕跡`, "折")|str_detect(`人為痕跡`, "敲"), "yes", "no"),
         burning =  ifelse(str_detect(`人為痕跡`, "火")|str_detect(`人為痕跡`, "燒"), "yes", "no")) %>%
  filter(is.na(refitted)) # for NISP
